---
title: 'Lesson 2: Working with data frames'
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
---

# **What is a data frame?**

A data frame is basically a type of table or spreadsheet where data is stored in columns and rows.

* In a data frame, all columns must be the same length. You cannot add a column that has fewer entries than the data frame has rows.
  + Exception - adding a column with one value will create a column where every entry is that value.
* All rows in a data frame must have a unique name. By default, these names are integers (e.g. 1, 2, 3, etc.), but can be in the form of text, as long as they are unique.

To make a dataframe from scratch, you can use the function `data.frame()`

For example, a simple data frame showing the results of a survey of 10 people which asked for their favorite ice cream flavor, whether they prefer a cone, and number of times per week they eat ice cream:

```{r}
dat <-  data.frame(flavor = c("vanilla", "chocolate", "strawberry", "vanilla", "vanilla",
                              "strawberry", "strawberry", "chocolate", "chocolate", "chocolate"),
                   cone = c(FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE),
                   count = c(3, 4, 1, 3, 2, 4, 4, 5, 2, 2))

dat

# the flavor row is being stored as a factor here, but I would prefer it to be character, so I will fix that here:
dat$flavor <- as.character(dat$flavor)
```

## Matrix vs. data frame

In R, both a matrix and a data frame are two dimensional arrays, but a matrix must contain values that are all of the same data type (numeric, character, factor, etc.), while in a data frame, each column can be of a different data type. In this way, data frames are more flexible.

However, in a given data frame column, all entries must be the same data type. If you try to add multiple data types to one column, R will coerce the data to one data type (usually character or factor).

## Converting matrix <-> data frame

You can convert a matrix to a data frame using `as.data.frame()` and from a data frame to a matrix using `as.matrix()`. This may be necessary sometimes, as certain functions only accept one or the other.

* e.g. Some heatmap functions only accept a matrix of numbers as input. So even if the data being used is coming from a larger data frame, the data for the data frame will need to be converted to a matrix using `as.matrix()`, either before the heatmap function or in the heatmap function itself.

```{r, eval = FALSE}
# For a data frame called df:
df.ht <- as.matrix(df[1:5, 1:5])

# OR
heatmap(as.matrix(df[1:5, 1:5]))
```

# **Reading data from a file**

R can read and write files containing data frames. This is especially helpful when you have very large data frames such as genomic datasets.

## File types

There are a few different file types you can get data frames from:

### CSV files

CSV, or comma separated value, files are commonly used to read and save data in R. In a CSV file, each line is a row of the data frame, with values for each column separated by commas (hence the name).

* If the CSV file has column and row names included, you will see a comma first in the first row, since there is nothing in that first space (row names do not get a column name).

### Excel and Google Sheet files

There are packages that allow you to read Excel (.xls, .xlsx) and Google Sheet (.gsheet) files directly into R. However, in my research I have not found one that is quick and easy. The good ones for Excel don't work with older .xls files, and Google Sheet can require authentication for sheets that aren't public (and that is a tedious process). For now, I would recommend exporting any Excel or Google Sheet files you want to use in R to CSV files.

In Excel, there are a few options for saving as a .csv. I always use the "CSV (Comma Delimited) (*.csv)" option and have never had any problems.

## Relative vs. absolute file locations

If you want to read data from any type of file in R, it's important to know about how R "sees" file locations. There are two ways you can tell R where a file is:

### Absolute location

This tells R *exactly* where the file is on your computer (or the internet), starting from the drive, e.g. `"C://Users/Andria/example/folder/with/a/file.csv"`.

I generally prefer to use this method, as it more foolproof, and you will only run into issues if the file itself moves, not the script or project.

### Relative location

This is relative to where you are currently located in the home directory. You can see what you current directory is by using `getwd()`, and change the working directory by using `setwd("C://Users/Andria/example/folder")`.

A relative location assumes that you are already in the working directory and goes from there, e.g. `"/with/a/file.csv"`. This can be useful at times, but to me is usually more complicated than it is worth.

## Tips for using files/folders in a script

This is an aside from today's main point, but it's related to good script writing.

In my scripts, I generally have a "set up" section at the top, which is shorter or longer depending on how complex the script is. If I will be accessing external files in the script, or writing files out, I usually set the locations up in their own section at the beginning. That way, if I need to run the script with a different set of data, I don't have to go searching throughout to look for file locations.

If I have just a few files to access or save, I will give each their own variable:

```{r, eval = FALSE}
datafile <- "C:/Users/Andria/example/file/input.csv"

outfile <- "C:/Users/Andria/example/file/output.csv"
```

If I have multiple files to access or write to, I may just save the folder as a variable. I do this more commonly with output, because I might generate multiple output files for each input, and then to run another input I just change one output folder and not every single file.

```{r}
outsave <- "C:/Users/Andria/example/outputfolder/"
```

Then, for each file that has to be written, I can generate the complete file name using `paste0()`, which will paste strings together with no space between them.

```{r}
paste0(outsave, "output.csv")
```

This saves space and keeps your code looking cleaner, as well as making your life easier when you have to run the same code 7 different times.

## Reading in CSV files

The function `read.csv()` can be used to read a .csv file into R as a data frame. It has a few arguments, the most important of which are:

* `file` - the location of the .csv file to be read in, as a character string (in quotes), as described above
* `header` - does the file contain column names? By default this is set to `TRUE`, which will read the first line as column names, but you can change it to `FALSE` if you don't have column names.
* `row.names` - this argument can tell R which column contains row names, if the file contains them, as a number (usually `1`). You could fix this the long way, but this argument makes it quick and easy.

Putting it all together, to read in a .csv file that has both row and column names:

```{r, eval = FALSE}
dat <- read.csv("C:/Users/Andria/example/outputfolder/input.csv", row.names = 1)
```

# **Writing data to a file**

Writing a CSV file from a data frame is very similar, and very simple, using the `read.csv()` function:

```{r, eval = FALSE}
write.csv(dat, "C:/Users/Andria/example/outputfolder/output.csv")
```


# **Accessing parts of a data frame**

There are several different ways you can access or see different parts of a data frame.

## Basic info

Data frames can get very big, and so calling the whole data frame can lead to a massive print out in your console that is not easy to read. But there are a few ways you can see what the data frame looks like without looking at the whole thing.

* `head(dat)` will give you just the first five rows of a data frame (or 5 entries in a vector, etc.). Similarly, `tail(dat)` gives the last five rows.
* `dim(dat)` will tell you the dimensions, or number of rows and columns of a data frame. You can get them individually by using `nrow(dat)` or `ncol(dat)`.
* `colnames(dat)` and `rownames(dat)` will give you the column names or row names of the data frame, respectively. This is helpful when I want to know where a particular column is in a very wide spreadsheet.

## Changing row/column names

You can change the column or row names using the same `colnames(dat)` and `rownames(dat)` function, and assigning a new set of values to them.

For example, let's label our rows with the names of our survey participants:

```{r}
rownames(dat) <- c("Arthur", "Bertha", "Cristobal", "Dolly", "Edouard", "Fay", "Gonzalo", "Hanna", "Isaias", "Josephine")

dat
```

## Accessing by row/column number

The rows and columns of a data frame can be accessed using square brackets `[]` after the data frame name. The row(s) to be accessed go in the brackets first, followed by a comma, and then the column(s) to be accessed.

For example:

```{r}
dat[2,1]
```

You can enter a range of numbers to be accessed:

```{r}
dat[1:5, 3]
```

Or put multiple numbers together using `c()`

```{r}
dat[c(1, 3), 1]
```

You can leave one side blank, meaning "all rows/columns."

```{r}
dat[,1]
```

You can put a minus sign in front of a number or set of numbers to mean "everything except these":

```{r}
dat[-2,]
```

## Accessing by row/column name

You can also use the row and column names to access the data. When possible, I really recommend doing this, as it will make your script more widely applicable. In a different dataset, the row and column orders may be different, and so you would have to re-enter all the correct numbers. But if the names of those rows and columns are the same, R will still find them.

If you want to get the results from our survey for a specific person:

```{r}
dat["Fay",]
```

Or you just want to see all results for the cone question:

```{r}
dat[,"cone"]
```

You can combine names just like numbers using `c()`:

```{r}
dat[c("Hanna", "Dolly"),]
```

Note that R returns the rows in the order that I asked for them, not the order they are in the original data frame.

Columns can also be accessed by name using a dollar sign (`$`). A data frame really stores each column as an individual vector, so this method pulls out that vector:

```{r}
dat$cone
```

This vector can also be subsetted by number like above, but since it is one dimensional, only one number or set of numbers is given in the square brackets:

```{r}
dat$cone[5]
```

## Changing data

Note that just like the `colnames()` and `rownames()` functions above, any of these methods for accessing the data can also be used to change the data. You can replace entire rows or columns, or individual entries.

Say, for example, that we realized we had written down one of our survey results incorrectly, and Cristobal's favorite ice cream flavor is not strawberry but chocolate:

```{r}
dat["Cristobal", "flavor"]

dat["Cristobal", "flavor"] <- "chocolate"

dat["Cristobal", "flavor"]
```

# **Filtering data frames by criteria**

There are times you may want to pull certain entries from your data frame based on a set of criteria.* For this, we will use functions like `which()` that let use logic expressions to tell R which parts we want to see.

* in fact, I recommend setting your script up to do this whenever possible, as it makes it easier to apply the same script to multiple inputs.

## Filtering by number

We can use greater than/less than/equal to logic to filter a data frame by values.

To do this, we'll use the `which()` function, which will return the numbers of the entries that meet the criteria. For example, if we want to know which people in our survey eat ice cream exactly 3 times per week, we can ask:

```{r}
which(dat$count == 3)
```

Notice that there are two equal signs in our logic statement. You must always use two symbols with an equal sign, but only one for less/greater than.

* `==` - "is equal to"
* `>=` - "is greater than or equal to"
* `<=` - "is less than or equal to"
* `>` - "is greater than"
* `<` - "is less than"
* `!=` - "is not equal to" - this will return everything that does not have the matching value

So, we can look for those who eat ice cream more than 4 times a week:

```{r}
which(dat$count >= 4)
```

But this only gives us the row number of these entries. To get the entry itself, we feed our `which()` function into the square brackets or column vector:

```{r}
dat[which(dat$count >= 4),]

# OR to ask for the flavor preference of these individuals:

dat$flavor[which(dat$count >= 4)]
```

We can also pull their names using the `rownames()` function, which can also be subsetted:

```{r}
rownames(dat)[which(dat$count >= 4)]
```

## Filtering by exact text

You can use text strings in a logic statement much like numbers, though for obvious reasons you can only use `==`. Say we want to pull the data for those people whose favorite flavor is strawberry:

```{r}
dat[which(dat$flavor == "strawberry"),]
```

### Filtering by text in a list

There is another method you can use to pull data by character strings in a list. The `%in%` operator tells you which entries from one list can be found in a second list.

Say you have a list of people from your suvey who are all from the same family, and you want to pull their data.

Here, I am going to save our list manually, but you could also import a list (such as a list of genes or a data frame with gene names) from a file.

```{r}
fam <- c("Arthur", "Dolly", "Fay", "Hanna", "Kayla")
```

To find out which rows match our family list:

```{r}
rownames(dat) %in% fam
```

Now this returns a `TRUE`/`FALSE` value for each of the ten rows, telling us which ones are in the `fam` list. We can feed this statement to the data frame call to get the actual data:

```{r}
dat[rownames(dat) %in% fam,]
```

Note that Kayla is in our family list but not our original survey data, so she is not shown here. If there is an entry in the second list that is not in the first, R will not throw an error, because it is only asking if the entries in the first list are in the last.

If you ever get confused by this, one thing that helps me get it straight is to look at the length of the output for each phrase:

```{r}
length(rownames(dat) %in% fam)

length(fam %in% rownames(dat))
```

## Filtering by TRUE/FALSE statements

You can also filter by logical `TRUE` or `FALSE` results in a similar way as text. You just use `TRUE` or `FALSE` with no quotes.

```{r}
dat[which(dat$cone == TRUE),]
```


## Filtering by incomplete/inexact text

With the above methods, you must be exact in the text that you use to query. If there is a typo or variation in the text, or the string is incomplete, the text will not be found.

That's where another method can come in handy. The `grep()` function matches text to a pattern, and it does not have to be an exact match, only a partial match (that is, the text you are searching needs to contain the pattern, but can contain other text around that pattern).

`grep()` primarily needs to arguments:

* the pattern to match
* the text to find the pattern in

in that order.

Say we want the rows of the individuals who have an "e" in their name:

```{r}
grep("e", rownames(dat))
```

Again, we can feed this function into our data frame:

```{r}
dat[grep("e", rownames(dat)),]
```

Notice that it didn't pull Edouard's row even though his name starts with E. That's because `grep()` is case-sensitive by default, but we can change this with the argument `ignore.case = TRUE`:

```{r}
dat[grep("e", rownames(dat), ignore.case = TRUE),]
```

The most common way I use `grep()` is to pull certain columns from a large data frame. For example, when I have DESeq2 output from multiple comparisons, each comparison has its own logFC and p value columns, labeled `somename.logFC` and `somename.padj`, respectively. For a heatmap, I only want the logFC values. Rather than type each column name individually, I will pull that data using `df[,grep("logFC", colnames(df))]`.

# **Other functions**

## Performing basic calculations

We can do some basic calculations on data frame columns, and save those results as a new column, if desired.

For example, say we want to take our data on the number of times each person eats ice cream in a given week and estimate the number of times they eat ice cream per month. We can multiply the week numbers by 4, and store this as a new column:

```{r}
dat$monthcount <- dat$count * 4

head(dat)
```

With this simple set up, R will take each entry in `dat$count` and multiply it by 4, then save that value as the corresponding entry for the new `dat$monthcount`.

We could also do a calculation using multiple columns:

```{r}
(dat$count + 1)/dat$monthcount
```

Again, this uses the corresponding values for `dat$count` and `dat$monthcount` to do a total of 10 calculations. This will also work for some simple functions.

However, some functions will not work like this. If we want to get the minimum value from two columns, we might try:

```{r}
min(dat$count, dat$monthcount)
```

But this only gives one value, because it is taking all of the values from `dat$count` and `dat$monthcount` and finding the minimum of all 20, not taking the minimum of each pair of two.

For these more complex types of calculations, we will need to use `for()` loops, which we can cover in a future lesson.

## Sorting

We can sort the data in a data frame by the row names or columns using the `order()`. This function takes a vector and returns the indices of the entries in order.

For example, we can get the order of rows from lowest `count` value to highest:

```{r}
order(dat$count)
```

Then we can feed this into the data frame. Since we are ordering by rows, we put the order in for the row numbers:

```{r}
dat[order(dat$count),]
```

Note that in ties, R defaults to the original ordering to break the tie.

By default, this orders things in ascending order (A to Z and 1, 2, 3, etc.), but the `decreasing` argument can be set to `TRUE` to reverse this.

For example:

```{r}
dat[order(rownames(dat), decreasing = TRUE),]
```

Orders the rows by reverse alphabetical order of the names.